---
title: "Data Vis example: homicide data"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
---

Load data from previous computations:
```{r getdat}
load("data/CA_homicide.RData")
```

One of the advantages of long format is that it allows us to use some of R's more powerful graphics tools such as the `ggplot2` and `lattice` packages (and it's what most statistics packages expect):

```{r ggplot1,message=FALSE}
library(ggplot2)
theme_set(theme_bw())  ## black-and-white theme
## set up basic plot:
p1 <- ggplot(mdat,aes(Year,Hom_rate,colour=Place))
```

Unlike base plots (which can only be saved/replayed through the RStudio interface), `ggplot` produces an R object which can then be printed (=displayed in a graphics window), or saved (or exported to a graphics file via `ggsave()`):

```{r ggplot2}
print(p1+geom_line())
```

We could add both lines and points:
```{r ggplot3}
print(p1+geom_line() +geom_point())
```

Might be better on a log scale, with a sensible y-axis label:
```{r ggplot4}
p1L <- (p1 + geom_line()+
        scale_y_log10()+
        labs(y="Homicides per 100,000 population")
)
print(p1L)
```

Maybe we don't care about time at all:
```{r boxplot}
b1 <- (ggplot(mdat,aes(x=Place,y=Hom_rate,
                      colour=Region))+
  geom_boxplot(outlier.colour=NULL)+  ## set outlier points to same colour
  scale_y_log10()+labs(y="Homicides per 100,000 population")
)
print(b1)
```

The labels are horrible.
We could rotate them 90 degrees to be vertical:
```{r vertlabels}
b1_vertlabels <- b1+theme(axis.text.x=element_text(angle=90))
print(b1_vertlabels)
```

In general if you want to tweak a `ggplot` plot, Google it or
search the [ggplot theme vignette](http://docs.ggplot2.org/dev/vignettes/themes.html) or the [ggplot cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) for more information ...

Rotating the whole plot is less familiar, but arguably better.
Here I'm also (1) changing the colour palette
and (2) changing the order of the `Place` variable, using `%+%` to
substitute a different set of data:

```{r coord_flip,message=FALSE,warning=FALSE}
library(dplyr)
mdat_sort <- mdat %>% mutate(Place=reorder(Place,Hom_rate))
print(b1 %+% mdat_sort + ## substitute sorted data
      coord_flip()+      ## rotate entire plot
      xlab("")+          ## x-label redundant
      scale_colour_brewer(palette="Dark2") ## change palette
      )
```

Maybe we want to make our line graph less busy: 
```{r facet_wrap}
print(p1L+facet_wrap(~Region))
``

We could also code population size by line width:
```{r plot3}
p2 <- ggplot(mdat,
             aes(Year,Hom_rate,colour=Region,size=log(Pop_2011),
                 group=Place))+geom_line(alpha=0.5)+
  scale_y_log10()+labs(y="Homicides per 100,000 population")
print(p2)
```

Using the [directlabels](http://directlabels.r-forge.r-project.org/) package (also on [GitHub](https://github.com/tdhock/directlabels)):

```{r directlabels,message=FALSE}
library(directlabels)
direct.label(p1L)+
  expand_limits(x=2004,y=0.1)
```

