---
title: "Canadian homicide rates"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
author: Ben Bolker
---

## Read data in wide format and play with it

Data are originally from [here](http://www.statcan.gc.ca/tables-tableaux/sum-som/l01/cst01/legal12b-eng.htm) (it might be possible to *scrape* this data set using the `XML` package ...).

I also got [data on population size in 2011 from Wikipedia](http://en.wikipedia.org/wiki/List_of_Canadian_provinces_and_territories_by_population) .

```{r popdat,echo=FALSE,eval=FALSE,message=FALSE}
## hidden stuff about 'scraping'
page <- "http://en.wikipedia.org/wiki/List_of_Canadian_provinces_and_territories_by_population"
library("XML")
popdat <- setNames(readHTMLTable(page)[[1]][,2:3],
                   c("Place","Pop_2011"))
## strip off junk and commas and convert to numeric
popdat[,2] <- as.numeric(gsub(",","",
                            gsub("^700[0-9]+000+","",popdat[,2])))
write.csv(popdat,file="popdat.csv",row.names=FALSE)
```
Read data.

```{r getdat}
dat <- read.csv("data/CA_homicide.csv",check.names=FALSE)
popdat <- read.csv("data/CA_popdat.csv")
```
We use `check.names=FALSE` to stop R from trying to sanitize the column names: this is reasonable if we plan to convert to long form and want to preserve the values (years, in this case).

These data are in wide format:
```{r showdat}
head(dat)
```

What if we want combine other information?
```{r regiondat}
head(rdat <- data.frame(Place=dat$Place,
      Region=c("all",rep("Atlantic",4),
             rep("East",2),
             rep("West",4),
             rep("North",3))))
```

Let's start by converting the data to long form (I've suppressed some warnings about "objects are masked"):
```{r gather,message=FALSE,warning=FALSE}
library("tidyr")
library("dplyr")
sdat <- dat %>%
    gather(year,homicides,-Place,convert=TRUE)
```
(we use `convert=TRUE` to convert the years back to numeric values)

Now combine all three data sets (`full_join` will automatically
match all columns with identical names across the data sets,
but it's better practice to specify the matching columns explicitly).

```{r join}
sdat2 <- sdat %>%
    full_join(rdat,by="Place") %>%
    full_join(popdat,by="Place")
```

If we just used the original data set (without the added stuff), it's fairly easy to get summary statistics by dropping the first row (so that we have a data frame that is all numeric) and computing means of rows and columns:
```{r sum1}
dmat <- dat[,-1]
rownames(dmat) <- dat[,1]
rowMeans(dmat)  ## means by place
colMeans(dmat)  ## means by year
```
(Don't forget the `na.rm` argument, unnecessary in this case, that can be provided to most R summary functions to get them to ignore `NA` values.)

If we want summary statistics from the full data set we can do

```{r,results="hide"}
sdat2 %>%
    group_by(Place) %>%
    summarise(mean=mean(homicides))
sdat2 %>%
    group_by(year) %>%
    summarise(mean=mean(homicides))
```

One more useful technique is reordering factors (representing categorical variables) in a sensible way.  Right now the 'places' (provinces, territories, etc.) are ordered alphabetically, R's default.

```{r}
sdat3 <- sdat2 %>%
    mutate(Place=reorder(Place,Pop_2011))
```
This will be useful in the future, but is different from the **order the data frame is stored in**, which we can modify via `arrange()` (use `desc(Pop_2011)` to arrange in descending order of population):

```{r}
sdat3 %>% arrange(desc(Pop_2011)) %>% head
```

I can also summarise by combinations of variables:
```{r}
sdat3 %>% group_by(year,Region) %>%
      summarise(mean=mean(homicides))
```

What if I want the mean and standard error?  R doesn't have a built-in "standard error of the mean" function so I define one on the fly:

```{r}
sem <- function(x) { sd(x)/sqrt(length(x)) }
region_avgs <- sdat3 %>% group_by(year,Region) %>%
    summarise(mean=mean(homicides,na.rm=TRUE),
              sem=sem(homicides))
```

What if I want to check the variables to see why they're `NA`?

```{r}
sdat3 %>% filter(year==2007 & Region=="all")
```

Sometimes it's useful to be able to go from long to wide format. `spread()` is the opposite of `gather()`: we specify a column in the current data set to spread out into new columns (`key`) and a column to use as the vales for the table (`value`)

```{r}
region_avgs %>% select(-sem) %>%
    spread(key=Region,value=mean)
```
In the long run it is generally easier to keep your data in long format and cast it to wide as necessary.

## tidyverse: summary

### advantages

- consistency among verbs
- piping format may be easier to read
- generally fast
- be like the cool kids

### disadvantages

- *non-standard evaluation* can be tricky/dangerous
- package-loading overhead
- annoy the old farts

