---
title: "Introduction to R"
author: Jonathan Dushoff and Ben Bolker
date:  26 April 2016
slide_level: 3
output: html_document
---

<!-- rendering
 HTML:
   rmarkdown::render("intro_Lecture_notes.rmd")
   rmarkdown::render("intro_Lecture_notes.rmd",output_format="ioslides_presentation")
   NOTE slide_level=3 doesn't do what I want! Makes section headers <3 into *section slides* ...

   rmarkdown::ioslides_presentation("intro_Lecture_notes.rmd",header_level=3)


https://www.rdocumentation.org/packages/rmarkdown/versions/1.3/topics/ioslides_presentation
   -->

```{r opts,echo=FALSE}
library(knitr)
opts_chunk$set(results="hide")  
```

<!-- BMB: not sure what this stuff is for?  Was it hidden in a previous version?  How is it helpful? -->

<!-- in markdown, two or more spaces at the end of a line specify a hard line break-->

# Basics

## R is a programming language

- You may or may not use R for programming tasks,  but it will help if you understand roughly how it works inside.
- What is a program?
    - A program is a collection of simple instructions (read, write, send, manipulate, calculate) that should be executed in a logical order
    - A program's flow is controlled by *control structures* (instructions about what instruction to do next)
    - What are some examples of programs, and how do they work?

## Calculation

R works as a calculator.

```{r arithmetic}
2+2
2439*41
259*3861
2+sqrt(92)
```

What about:

```{r precedence}
7+3*9
32/4*4
```
Figure out what the second one does, and
then explain how you would write it more clearly.

## Assigning values

R can remember numbers and give them back to you. This is the foundation
of being able to write programs that do many things in a specified
order.

```{r assign}
x <- 23
print(x)
x + 17
y <- x + 17
y
```

# Objects

----------------------------------------------------------------------


R keeps track of a variety of objects: data structures and functions
mostly. You can see all of your user-defined objects by typing
`objects()` (the parentheses mean that you are calling a *function*;
more on this later).

## Values

In addition to numbers, R has other kinds of values (also known as "atomic types"). The main ones we're interested in are `character`, `numeric` and `logical` (ie., `TRUE` or `FALSE`). `str()` is one way to try to figure out what your R object is.


```{r values}
num <- 3
str(num) 
 
char <- "Hello, class?"
str(char)
 
logic <- TRUE ## Note lack of quotes
str(logic)
```

## Other types

R objects are divided into *types*. Other than the above, we're mostly interested in three of these: *vectors*, *lists* and *functions*.

## Vectors

A vector consists of zero or more values of the same *storage mode*:

```{r vector}
words <- c("Mary", "had", "a", "little", "lamb")
# c() puts elements together into a vector
str(words)
 
print(v <- 1:5)  ## 'm:n' creates a sequence from m to n
str(v)
```

## Vector math

R does math on vectors directly.

<!-- this flavor of markdown / pandoc needs an empty line before an itemized list -->

* If we add (for example) two vectors we add each pair
of corresponding elements.
* If we multiply a vector by a *scalar* (a
single number) we multiply each element by the scalar.
* Avoid arithmetic involving two vectors of different length (except the scalar case, above)

```{r vector_math}
v <- 1:5
w <- c(0, 1, 1, 2, 4)
v+w
2*w
```


## No scalars

R does not have any particular idea of individual numbers (or strings,
etc.); you represent a number as a vector of length 1. This has
advantages and disadvantages.



## Lists

A list is a bunch of things in order. Those things can be any R object - vectors, functions, other lists ...

```{r list}
L <- list(1:3, "Apple tree", TRUE)
str(L)
str(L[1]) ## L[1] picks out the first element of the list
LL <- list(L, c(2, 7, 9))
str(LL)
str(LL[1])
```

## Functions

Functions are "called" using parentheses. A function is a set of
commands that uses whatever "arguments" are inside the parentheses and
does things (including possibly "returning" a result).

What functions have we seen so far?

Example: `mean()` in its simplest form takes a vector argument and
returns the mean.

```{r mean}
mean(c(2, 5, 11))
 
x <- 1:10
m_x <- mean(x)
print(m_x)
```

## About functions

You can learn about any built-in function by using R's help: type
`?"mean"` or `help("mean")`.

Arguments can be passed to functions in order, or by using names. For
example, `mean` takes an optional argument `trim`.

```{r trim}
x <- (1:10)^2
mean(x)
mean(x, trim=0.1)
```

## Function syntax

`=` is a synonym for `<-` at the top level (and many people use it).
Be aware that they are not synonyms inside a function call: `<-`
still means assignment, but `=` is used to name an argument.

## Variable names

R variable names are case-sensitive. This means that `m` and `M` are
different variables. In general, you should not try to take advantage of this, because it may cause confusion.

R variable names have to start with a letter, and certain characters
(particularly space characters) are not allowed. Good naming conventions
include:

-   Using mostly letters, underscores (`_`) and dots (`.`)
-  Not using potentially confusing names like `l` or `O`.
-  Not using built-in names (like `c`, `list`, or `data`) for variables.
-   Make readable variable names using `camelCase`, `snake_case`, or `kebab.case`
-   Avoid `variableNamesThatAreExcessivelyLong`

# Compound objects

----------------------------------------------------------------------

R is a sophisticated system that builds complicated structures on top of
these simple objects. Hopefully we'll develop an intuition for more of
that as we go along. People interested in nuts and bolts can also look
at the (rather scary!) R language definition on the [R manuals
page](http://cran.r-project.org/manuals.html). For now, we'll talk about
just a few structures.

## Matrices

In R, matrices are basically vectors that understand that they have a
shape. R knows how to do matrix multiplication on them, which is very
cool, and other kinds of linear algebra. There are many ways to make a
matrix: use `?"matrix"` to get more information.

```{r matrix}
m <- matrix(c(1, 3, 0, 1), nrow=2)
print(m)
# R arranges vectors into matrices in COLUMN-FIRST order, by default
 
# Elementwise ("vector") multiplication (or "Hadamard product")
m*m
 
# Matrix multiplication
m %*% m
```

## Data frames

Data frames are lists that have a rectangular shape (they look like, and
can be used a like matrices, but don't have to be of a single "mode" -
in particular you can mix numbers, factors (see below), dates, ...).

We'll talk more about data frames when we start dealing with data.

## Factors

Factors are used to describe items that can have a known set of values
(gender, social class, etc.) - *categorical variables* in statistical
terms. We will also talk more about them later.

# More stuff

## Control structures

Control structures in R include:

-  loops: we will use `for`, `lapply` and `apply`, among others.
    `lapply` stands for **l**ist **apply**.
-   conditionals: `if` and `ifelse`. Use `ifelse` with care!

```{r control}
v <- 1:10
for (x in v) {
  print(x^2)
}
for (x in v){
    if(x>=4) {
        print(x^2)
    }
}
```

## Control syntax

R uses `==` to test equality, and `!=` to test inequality. (R will often
guess and warn you if you wrongly try to use `=`, but not always ...)

* Use `&` for "and", and `|` for "or" when testing vectors in parallel
	* `&&` and `||` test single conditions; it's usually OK to stick to `&` and `|`.

## Syntactic principles

You may have noticed that we sometimes just state something in order to
print its value. It's better
practice to always say `print`, when that's what you want.
* Clearer, and less fragile

Similarly, you usually don't need to say `return` to return a function value … but you should. 

Clarity and explicitness are worth more than saving a few keystrokes.

* When R was invented, this was less clear

## Names

R's primary objects often have associated "attributes". The information
that tells R what shape a matrix (or higher-dimensional array) is is an
attribute. R also has "name" attributes: `names`, `rownames` and
`colnames`.

You can select single elements from an R object using the element
operator `[[]]`, and subsets using the subset operator `[]`. If
the object has dimensions, you can separate them with a comma.

## Selecting

You can also select things by name, using the syntax `v["name"]` or
`v[n]`, if `n` is a variable that contains the name. In general it's a
good idea to use names instead of numbers whenever you can, because (1)
your code will be easier to read (`x["temperature"]` instead of `x[21]`)
and (2) your code will be more robust (i.e., if you modify something so
that temperature is now in position 22 rather than 21).

## Selection example

```{r indexing}
# InsectSprays is a built-in data set in R
class(InsectSprays)
head(InsectSprays)
names(InsectSprays)
 
# Select the first list element of InsectSprays
InsectSprays[[1]]
 
# Select a subset of InsectSprays containing only the first list element
InsectSprays[1]
 
# Select the first column of InsectSprays
InsectSprays[, 1]
 
# Select part of the first column of InsectSprays
InsectSprays[1:10, 1]
 
# Select the first list element of InsectSprays by name
InsectSprays[["count"]]
 
# Use a convenient but sometimes dangerous short-cut for this
InsectSprays$count
```

## Arcane

You may have been surprised by the column-selection results. We used the
subset operator, but got a single element instead. R noticed we only
selected one thing, and "collapsed" the object for our "convenience".
This sort of thing is a common problem in R, probably due to its rather
ancient history, and the tension between structure and convenience
(particularly convenience for interactive use).

In programs, it is bad practice to let R do this (because it will
confuse the program when the structure of your results changes). In this
case, you can avoid it by selecting from the list.

## Being explicit

It's unfortunate that R makes you do stuff like this…

```{r arcane}
# A data frame
InsectSprays[, 1:2]
#  An inappropriate "collapse" to a simpler form
InsectSprays[, 1]
# Avoid inappropriate collapse
InsectSprays[, 1, drop=FALSE]
```
