---
title: "Intro to tidyverse"
author: "Jonathan Dushoff and Ben Bolker"
date:  "`r format(Sys.time(),'%d %B %Y')`"
slide_level: 3
output: html_document
---


```{r opts,echo=FALSE}
library(knitr)
opts_chunk$set(results="hide")  
```

# The tidyverse

## basic concepts

- a set of R packages: https://www.tidyverse.org/
- advantages
    - expressiveness
    - speed
	- new hotness
- disadvantages
    - incompatibilities with base R
	- rapid evolution
	- non-standard evaluation

## Sample data

We want to get some data. Before executing the code below:

- think about where you want to keep your stuff for this class
- set your **working directory** accordingly (`setwd()`, or in RStudio `Session / Set Working Directory / Choose Directory`)
- create a `data/` subdirectory if you want, or edit it out of the `destfile` argument below

```{r eval=FALSE}
download.file(url="https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```

## And now ...

We will continue with the [Data Carpentry](https://datacarpentry.org/) [lesson for R in ecology](https://datacarpentry.org/R-ecology-lesson/):

https://datacarpentry.org/R-ecology-lesson/03-dplyr.html
