---
title: "Linear model testing"
---

## Variable-level P values

- `anova(model1,model2)` compares two *nested* models
- `car::Anova` and `drop1` are wrappers that allow you to do many simple anovas at once (one for each variable (or term) that R thinks it should drop).

When you are *estimating* variance (e.g., with a normal response variable) you should use an $F$ test.

When you can *calculate* variance from your assumptions (e.g., with a Poisson or binomial), you should use a $\chi^2$ test.

## Simple model

```{r tests1}
load("data/contrasts2.RData")
print(summary(lmTL1))
lmlight <- lm(grahami~light, data=lizards)
lmtime <- lm(grahami~time, data=lizards)
## Test time by comparing with light model (dropping time)
anova(lmTL1, lmlight, test="F")
## Test light (should be exactly the same test as in summary)
anova(lmTL1, lmtime, test="F")
## In this simple case, drop1 and car::Anova
##  will do exactly the same thing we did above
drop1(lmTL1, test="F")
car::Anova(lmTL1)
```

## Interaction model

```{r test2}
print(summary(lmTL2))
## Test time by comparing with light model (drop time AND interaction)
anova(lmTL2, lmlight, test="F")
## Test light
anova(lmTL2, lmtime, test="F")
## drop1 now tests the whole interaction, also very useful
drop1(lmTL2, test="F")
car::Anova(lmTL2)  ## 'type II'; drop interaction before testing
car::Anova(lmTL2, type='III')  ## 'type III'; test time at baseline light & v. versa
```

## Multiple comparisons

Some notes on the theory behind multiple comparisons.

Multiple comparisons are tricky, and you probably don't need to do them.  Unlike variable-level tests, they will depend on how you parameterize your model.  Your results for a given multi-level variable can even depend on how you parameterize other variables.

```{r multcomp,message=FALSE}
library(multcomp)
print(mc <- glht(lmTL1, linfct=mcp(time="Tukey")))
cld(mc)
```

```{r emmeans,message=FALSE}
library(emmeans)
(e1 <- emmeans(lmTL1, "time"))
pairs(e1)
```

