---
title: "Mixed model examples"
---

## Examples

- Most examples here taken from [Vincent Zoonekynd's page](http://zoonek2.free.fr/UNIX/48_R/14.html)
- See also [supplementary material from Bolker book chapter](http://bbolker.github.io/mixedmodels-misc/ecostats_chap.html) (source code [here](https://github.com/bbolker/mixedmodels-misc/blob/master/ecostats_chap.rmd))

## Formulae

Formula  |                  Meaning
-------------------------|------------------------------------
y ~ x                    |     No random effects
y ~ x + (1|g)            |   The intercept is a random effect

```
y ~ x + (1|site/block)      Nested random effects (block within site)
y ~ x + (1|site) + (1|year) Crossed random effects
y ~ x + (1|site:block)      Interaction (only block within site)
y ~ x + (x|g)               Intercept and slope are random effects
y ~ (x|g)                   Zero slope on average (weird!)
y ~ x + (1|g)+(0+x|g)       Independent slope and intercept
```

## Basic look at the data

- Sometimes called a *spaghetti plot*
- Alternative: use `+facet_wrap(~Subject)`

```{r mmex1}
library(lme4)
library(ggplot2); theme_set(theme_bw())
q0 <- (ggplot(sleepstudy, aes(Days, Reaction, colour = Subject))
    + geom_point())  ## points only, use later
print(q0+geom_line())
```

## Basic model fits

``` {r mmex2,message=FALSE}
library(nlme)
lm1 <- lmList(Reaction~Days|Subject,data=sleepstudy) ## per-group fit (fixed)
lm2 <- lme(Reaction~Days,random=~1|Subject,data=sleepstudy) ## rand intercept
lm3 <- lme(Reaction~Days,random=~Days|Subject,data=sleepstudy) ## rand slopes
```

## Compute predictions

```{r preds}
pp <- expand.grid(Days=0:9,Subject=levels(sleepstudy$Subject))
pp1 <- cbind(pp,Reaction=predict(lm1,newdata=pp))
pp2 <- cbind(pp,Reaction=predict(lm2,newdata=pp))
pp3 <- cbind(pp,Reaction=predict(lm3,newdata=pp))
summary(lm3)
```

## plot predictions

``` {r mmex3, echo=FALSE}
print(q0
      + geom_line(data=pp2)
      + geom_line(data=pp3)
      + geom_line(data=pp1,lty=2))
```

## the same thing with lme4::lmer

``` {r mmex4, echo=FALSE}
library(lme4)
lm4 <- lmer(Reaction~Days+(Days|Subject),data=sleepstudy)
summary(lm4)
```
