---
title: "Reshaping data with `tidyr`"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
author: Ben Bolker
------

```{r pkgs}
library(tidyr)
```

Look at some example data that comes with the package:

```{r view}
smiths
```

The default `gather()` operation squashes everything too far,
including the subject name and time in the value column ...

```{r gather1}
gather(smiths)
```

We can specify that we only want to gather the `age` and `weight` variables
(however, we have to specify the name of key and value columns explicitly).
```{r gather2}
print(smelt <- gather(smiths, key="var", value="value",
       c(age,weight)))
```

Alternatively we could specify that we
want to gather everything *but* the subject name and time variable:
```{r gather3}
gather(smiths, key="var", value="value",
       -c(subject,time))
```

Make a column for each subject (= a row for each measurement)

```{r col_per_subj}
spread(smelt, key=subject, value)
```

Make a column for each value (= a row for each person):

```{r col_per_value}
spread(smelt, key=var, value)
```

Take the mean for each variable:

```{r aggregate1,message=FALSE}
library(dplyr)
smelt %>% group_by(var) %>% summarise(mean=mean(value))
```

Insist that R calculate a mean:

```{r na1}
smelt %>% group_by(var) %>% summarise(mean=mean(value,na.rm=TRUE))
```
Insist that R calculate a mean, but also report how many values are in each mean:

```{r mean_n1}
smelt %>% group_by(var) %>% 
    summarise(mean=mean(value,na.rm=TRUE),
              n=length(na.omit(value)))
```

If you're careful you can `na.omit()` first (be careful, this will
remove all rows with *any* `NA` values):
```{r mean_n2}
smelt %>% na.omit %>% group_by(var) %>% 
    summarise(mean=mean(value,na.rm=TRUE),
              n=n())
```
