---
title: "Canadian homicide rates"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
author: Ben Bolker
---

```{r setup, include=FALSE}
##  bmb needs this to get transparency rendered properly
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
```

This is a joint example of (1) rearranging data and (2) graphing data.

# Rearranging data

## Read data in wide format and play with it

Data are originally from [here](http://www.statcan.gc.ca/tables-tableaux/sum-som/l01/cst01/legal12b-eng.htm) (it might be possible to *scrape* this data set using the `XML` package ...).

I also got [data on population size in 2011 from Wikipedia](http://en.wikipedia.org/wiki/List_of_Canadian_provinces_and_territories_by_population) .

```{r popdat,echo=FALSE,eval=FALSE,message=FALSE}
## hidden stuff about 'scraping'
page <- "http://en.wikipedia.org/wiki/List_of_Canadian_provinces_and_territories_by_population"
library("XML")
popdat <- setNames(readHTMLTable(page)[[1]][,2:3],
                   c("Place","Pop_2011"))
## strip off junk and commas and convert to numeric
popdat[,2] <- as.numeric(gsub(",","",
                            gsub("^700[0-9]+000+","",popdat[,2])))
write.csv(popdat,file="popdat.csv",row.names=FALSE)
```
Read data (from two separate files).

```{r getdat, message=FALSE}
library(tidyverse)
## using col_types=cols() suppresses the messages about which type is which
dat <- read_csv("../data/CA_homicide.csv", col_types=cols())
popdat <- read_csv("../data/CA_popdat.csv", col_types=cols())
```

These data are in wide format:
```{r showdat}
head(dat)
```

Note that `read_csv()` has **not** modified the column names to make
them legal R variables names (`read.csv()` will do that by default:
use `check.names=FALSE` to suppress this)

What if we want combine other information?
```{r regiondat}
rdat <- tibble(Place=dat$Place,
      Region=c("all",rep("Atlantic",4),
               rep("East",2),
               rep("West",4),
               rep("North",3)))
head(rdat)
```

Let's start by converting the data to long form:
```{r gather}
sdat <- (dat
    %>% pivot_longer(names_to="year",values_to="homicides",-Place,
                     names_transform=list(year=as.numeric))
)
head(sdat)
```
(we use `names_transform` to convert the years back to numeric values)

Now combine all three data sets (`full_join` will automatically
match all columns with identical names across the data sets,
but it's better practice to specify the matching columns explicitly).

```{r join}
sdat2 <- sdat %>%
    full_join(rdat,by="Place") %>%
    full_join(popdat,by="Place")
```

If we just used the original data set (without the added stuff), it's fairly easy to get summary statistics by dropping the first row (so that we have a data frame that is all numeric) and computing means of rows and columns:
```{r sum1}
dmat <- as.matrix(dat[,-1])
rownames(dmat) <- dat$Place
rowMeans(dmat)  ## means by place
colMeans(dmat)  ## means by year
```
(Don't forget the `na.rm` argument, unnecessary in this case, that can be provided to most R summary functions to get them to ignore `NA` values.)

If we want summary statistics from the full data set we can do

```{r fullsum,results="hide"}
sdat2 %>%
    group_by(Place) %>%
    summarise(homicides=mean(homicides))
## fancier
sdat2 %>%
    group_by(year) %>%
    summarise(across(homicides,list(mean=mean,sd=sd)))
```

One more useful technique is reordering factors (representing categorical variables) in a sensible way.  Right now the 'places' (provinces, territories, etc.) are ordered alphabetically, R's default.

```{r reorder}
sdat3 <- sdat2 %>%
    mutate(Place=forcats::fct_reorder(Place,Pop_2011))
## equivalent
sdat3 <- sdat2 %>%
    mutate(across(Place,~forcats::fct_reorder(.,Pop_2011)))
```
This will be useful in the future, but is different from the **order the data frame is stored in**, which we can modify via `arrange()` (use `desc(Pop_2011)` to arrange in descending order of population):

```{r arrange}
(sdat3
    %>% arrange(desc(Pop_2011))
    %>% head()
)
```

I can also summarise by combinations of variables:
```{r sumcomb}
sdat3 %>% group_by(year,Region) %>%
    summarise(across(homicides,mean),.groups="drop")
```

What if I want the mean and standard error?  R doesn't have a built-in "standard error of the mean" function so I define one when I need it:

```{r meanse}
sem <- function(x) { sd(x)/sqrt(length(x)) }
region_avgs <- sdat3 %>% group_by(year,Region) %>%
    summarise(across(homicides,list(mean=~mean(.,na.rm=TRUE),
                                    sem=sem)),
              .groups="drop")
head(region_avgs)
```

**Question: why do I have NA values?**

Drilling down to check some values:

```{r check}
sdat3 %>% filter(year==2007 & Region=="all")
```

Sometimes it's useful to be able to go from long to wide format. `pivot_wider()` is the opposite of `pivot_longer()`: we specify a column in the current data set to spread out into new columns (`key`) and a column to use as the vales for the table (`value`)

```{r}
(region_avgs
    %>% select(-homicides_sem)
    %>% pivot_wider(names_from=Region,values_from=homicides_mean)
)
```


Save the results:

```{r save}
saveRDS(sdat3,file="../data/CA_homicide.rds")
```

For analysis in R, it is generally best to keep your data in long format and `pivot_wider()` it as necessary (e.g. when creating a human-readable table for output).

# Pictures

Load data from previous computations (can also get it [here](../data/CA_homicide.rds)
```{r getdat2}
mdat <- readRDS("../data/CA_homicide.rds")
```

One of the advantages of long format is that it allows us to use some of R's more powerful graphics tools such as the `ggplot2` and `lattice` packages (and it's what most statistics packages expect):

```{r ggplot1,message=FALSE}
library(ggplot2)
theme_set(theme_bw())  ## black-and-white theme
## set up basic plot:
p1 <- ggplot(mdat,aes(year,homicides,colour=Place))
```

Unlike base plots (which can only be saved/replayed through the RStudio interface), `ggplot` produces an R object which can then be printed (=displayed in a graphics window), or saved (or exported to a graphics file via `ggsave()`):

```{r ggplot2}
print(p1+geom_line())
```

We could add both lines and points:
```{r ggplot3}
print(p1+geom_line() +geom_point())
```

Might be better on a log scale, with a sensible y-axis label:
```{r ggplot4}
p1L <- (p1
    + geom_line()
    + scale_y_log10()
    + labs(y="Homicides per 100,000 population")
)
print(p1L)
```

(Zero values get squashed against the lower axis)

Maybe we don't care about time at all:
```{r boxplot}
b1 <- (ggplot(mdat,aes(x=Place,y=homicides,
                       colour=Region))
    + geom_boxplot(outlier.colour=NULL)  ## set outlier points to same colour as boxes
    + scale_y_log10()
    + labs(y="Homicides per 100,000 population")
)
print(b1)
```

The x-axis tick labels overlap enough to be unreadable (unless we resize the
plot to be ridiculously long and narrow).

We could rotate them 90 degrees to be vertical:
```{r vertlabels}
b1_vertlabels <- b1+theme(axis.text.x=element_text(angle=90))
print(b1_vertlabels)
```


In general if you want to tweak a `ggplot` plot, Google it or
search the [ggplot theme documentation](https://ggplot2.tidyverse.org/reference/theme.html) or the [ggplot cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) for more information ...

Rotating the whole plot is less familiar, but arguably better.
Here I'm also (1) changing the colour palette
and (2) changing the order of the `Place` variable, using `%+%` to
substitute a different set of data into an existing plot:

```{r coord_flip,message=FALSE,warning=FALSE}
mdat_sort <- mdat %>% mutate(across(Place,~forcats::fct_reorder(.,homicides)))
print(b1
      %+% mdat_sort  ## substitute sorted data
      + coord_flip()      ## rotate entire plot
      + xlab("")          ## x-label redundant
      + scale_colour_brewer(palette="Dark2") ## change palette
      )
```

Maybe we want to make our line graph less busy: 
```{r facet_wrap,warning=FALSE}
print(p1L+facet_wrap(~Region))
```

We could also code population size by line width:
```{r plot3, warning=FALSE}
p2 <- (ggplot(mdat,
             aes(year,homicides,colour=Region,size=Pop_2011,
                 group=Place))
    + geom_line(alpha=0.5)
    + scale_y_log10()
    + scale_size_continuous(trans="log10")
    + labs(y="Homicides per 100,000 population")
)
print(p2)
```

Using the [directlabels](http://directlabels.r-forge.r-project.org/) package:

```{r directlabels,message=FALSE, warning=FALSE}
library(directlabels)
print(p1L
    + expand_limits(x=2014)  ## add a little more space
    + geom_dl(aes(label=Place),method="last.bumpup") 
    + theme(legend.position="none")  ## don't need the legend any more
)
```

We'd have to work a little harder to avoid clipping the "Yukon" label ...

