---
title: "Mixed model examples"
output:
  html_document:
    code_folding: hide
bibliography: ../qmee.bib	
author: Ben Bolker
---

## Load packages ...

```{r pkgs,message=FALSE}
## fitting
library(lme4)
library(glmmTMB)
library(lattice)
## diagnostics etc.
library(broom.mixed)
library(DHARMa)
library(performance)
library(sjPlot)
library(dotwhisker)
##
library(ggplot2); theme_set(theme_bw())
library(cowplot)
```

## Examples

- Most examples here taken from [Vincent Zoonekynd's page](http://zoonek2.free.fr/UNIX/48_R/14.html)
- See also [supplementary material from Bolker book chapter](http://bbolker.github.io/mixedmodels-misc/ecostats_chap.html) (source code [here](https://github.com/bbolker/mixedmodels-misc/blob/master/ecostats_chap.rmd))

## Formulae {.smaller}

Formula  |                  Meaning
-------------------------|------------------------------------
`y ~ x`                    |  No random effects
`y ~ x + (1|g)`            |  The intercept is a random effect
`y ~ x + (1|site/block)`   |   Nested random effects (block within site)
`y ~ x + (1|site) + (1|year)` | Crossed random effects
`y ~ x + (1|site:block)`     | Interaction (only block within site)
`y ~ x + site + (1|site:block)` | Fixed site, random block within site
`y ~ x + (x|g)`              | Intercept and slope are random effects
`y ~ (x|g)`                  | Zero slope on average (weird!)
`y ~ x + (1|g)+(0+x|g)`     | Independent slope and intercept


## Look at the data

- *Spaghetti plot*; alternatively `+facet_wrap(~Subject)`

```{r mmex1}
print((q0 <- ggplot(sleepstudy, aes(Days, Reaction, colour = Subject))
    + geom_point()) + geom_line())
```

(Alternatives; connect, don't colour; suppress legend; separate individuals into separate facets ... for scatterplots, try `ggalt::geom_encircle()`. See examples [here](https://bbolker.github.io/goettingen_2019/notes/datavis.html) as well.)

## Basic model fits

```{r mmex2,message=FALSE}
library(lme4)
## per-group fit (fixed)
lm1 <- lmList(Reaction~Days|Subject, data=sleepstudy)
## fixed effect
lm1B <- lm(Reaction~Days*Subject, data=sleepstudy)
## random intercept
lm2 <- lmer(Reaction~Days+(1|Subject), data=sleepstudy)
## random slopes
lm3 <- lmer(Reaction~Days+(1+Days|Subject), data=sleepstudy) 
```

Use *random slopes* when feasible, i.e. when some individuals are measured on more than 1 day [@schielzeth_conclusions_2009]

## Compute predictions

```{r preds}
pp <- expand.grid(Days=0:9, Subject=levels(sleepstudy$Subject))
pp <- (pp                  
  |> cbind(pred_int = predict(lm2, newdata=pp))
  |> cbind(pred_int_slope = predict(lm3, newdata=pp))
  |> cbind(pred_fixed = predict(lm1B, newdata=pp))
  |> tidyr::pivot_longer(starts_with("pred"))
)
```

## plot predictions

```{r mmex3, fig.width=10, out.width = "100%"}
theme_set(theme_classic()+theme(legend.position="none"))
q0+geom_line(data=pp, aes(y=value)) + facet_wrap(~name)
```

## diagnostics

* `DHARMa::simulateResiduals()`
* `performance::check_model()`

```{r, echo = FALSE}
performance::check_model(lm3, check = "reqq")
```

## coefficient plots

Use `broom.mixed` (in conjunction with `dotwhisker::dwplot`)

- use `tidy` directly for more control (`?tidy.merMod`)
- may want to specify `effects="fixed"`, `conf.int=TRUE`, `conf.method="profile"` (for CIs on random effects parameters)

```{r coefplots, results="hide", fig.keep="none", cache = TRUE, message = FALSE}
tt <- tidy(lm3, conf.int=TRUE, conf.method="profile")
```

##

```{r re_dotplot, warning=FALSE}
dwplot(tt)
```

## random effects plots

```{r replot}
lattice::dotplot(ranef(lm3))
```

Or `sjPlot::plot_model()`

## post-hoc stuff

`emmeans`, `effects`, `car::Anova()`, `drop1`, etc. all work ...

## model choice considerations

- what is the **maximal model**?
- Which effects vary *within* which groups?
- If effects don't vary within groups, then we *can't* estimate among-group variation in the effect
     - convenient
     - maybe less powerful (among-group variation is lumped into residual variation)
- e.g. female rats exposed to different doses of radiation, multiple pups per mother, multiple measurements per pup (labeled by time). Maximal model ... ?

---

Maximal model **often won't actually work**

e.g.

- *Culcita* (coral-reef) example: randomized-block design, so each treatment (none/crabs/shrimp/both) is repeated in every block; thus `(treat|block)` is maximal
- CBPP data: each herd is measured in every period, so in principle we could use `(period|herd)`, not just `(1|herd)`

## Random-slopes models

- what does `(x|g)` (or `(1+x|g)` really do?
- both intercept (baseline) and slope vary across groups
- estimates **bivariate** zero-centered distribution:

$$
(\textrm{intercept}, \textrm{slope}) =
\textrm{MVN}\left(\boldsymbol 0,
\left[
\begin{array}{cc}
\sigma^2_{\textrm{int}} &
\sigma_{\textrm{int},\textrm{slope}} \\
\sigma_{\textrm{int},\textrm{slope}} & 
\sigma^2_{\textrm{slope}}
\end{array}
\right]
\right)
$$

## `equatiomatic` package

```{r equatiomatic, results="asis"}
equatiomatic::extract_eq(lm3)
```

## What is a practical model?

- Avoid **singular** fits
- singular = zero variances, +/- 1 correlations
- More subtle for larger models: use `isSingular()`

## How to simplify?

- remove random effects
- simplify random effects [@bates_parsimonious_2015;@matuschek_balancing_2017;@barr_random_2013]
   - drop slopes
   - force correlations to zero (use `1+x||g` to fit intercept and slope independently)
   - structured random effects (diagonal, compound symmetric ...)
   
## Convergence failures 

- convergence failures are common
- what do they really mean? how to fix them? when can they be ignored?
- **approximate** test that gradient=0 and curvature is correct
- scale and center predictors; simplify model
- use `?allFit` to see whether different optimizers give sufficiently similar answers
     - `$fixef`, etc.: are answers sufficiently similar?
     - `$llik`: how similar is goodness-of-fit?

## allFit

```{r allFit_ex, cache=TRUE, message=FALSE}
m1 <- lmer(Reaction ~ Days + (Days|Subject), sleepstudy)
aa <- allFit(m1)
ss <- summary(aa)
names(ss)
ss$fixef
```

## allFit plot

```{r fig.width = 12, out.width = "100%"}
source("https://raw.githubusercontent.com/lme4/lme4/refs/heads/master/misc/plot_allFit.R")
par(las=1); tpar(fmar = c(1,12,1,1))
plot(aa, flip = TRUE)
```

## GLMMs: integration techniques

- PQL, Laplace, adaptive Gauss-Hermite quadrature (AGHQ)
- fast/flexible/inaccurate $\to$ slow/inflexible/accurate
- **amount of data per group** is what matters
    - number of points, info per point (binary vs count)
- set `nAGQ=` to a larger value (1=Laplace; 25 should be more than enough)

## GLMMs: more details

- `glmer` pretty reliable
- `glmmTMB` more flexible, faster on extended models (NB etc.)

## Other resources

- [Mixed Models task view]()
- the [GLMM FAQ](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html)

## References
