---
title: "Ordinal models"
output:
  html_document:
    code_folding: hide
bibliography: ../qmee.bib	
author: Ben Bolker
---

## Load packages ...

```{r pkgs,message=FALSE}
## fitting
library(MASS) ## polr
library(ordinal)
library(tidyverse); theme_set(theme_bw())
## Bayesian solutions
library(MCMCglmm)
library(brms)
library(patchwork)
## remotes::install_github("https://github.com/bbolker/ordinalsim")
library(ordinalsim)
library(sure)
```

## Goal

* estimate effects of categorical and continuous predictors when the response is *ordinal*
* categorical but ordered; we don't know distances between successive levels
* CLM = **cumulative link** model
* 

## house data

"Frequency table from a Copenhagen housing conditions survey"


```{r}
## FIXME: compute weighted means
pd <- position_dodge(width = 0.25)
ggplot(housing, aes(x=Type, y = as.numeric(Sat), colour = Infl)) +
  geom_point(aes(size=Freq), position = pd) +
  facet_wrap(~Cont, labeller = label_both) +
  scale_y_continuous(breaks = 1:3,
                     labels = levels(housing$Sat)) +
  labs(y = "Satisfaction")
```

```{r}
house.plr <- polr(Sat ~ Infl + Type + Cont, weights = Freq, data = housing)
print(ss <- summary(house.plr))
```

Understanding the intercept terms:
```{r}
## select baseline predictors (first level of each factor)
get_pframe <- function(Infl_sel = "Low",
                       Type_sel = "Tower",
                       Cont_sel = "Low") {
  subset(housing,
                 Infl == Infl_sel &
                 Type == Type_sel &
                 Cont == Cont_sel,
                 select = -c(Sat, Freq))[1,]
}

pframe <- get_pframe()
## predict probs from fitted model
(probs <- predict(house.plr, newdata = pframe, type = "probs"))

zeta <- ss$zeta

## hand calculation
c(plogis(zeta[1]), diff(plogis(zeta)), 1-plogis(zeta[2]))
```

@gambarotaOrdinal2024

```{r}
cat_latent_plot(prob = probs)
```

## Assumptions

* independence
* **proportional odds** or equivalent: changes to predictors (change in category, or a change in a numeric variable) shifts log-odds of neighbouring categories (Medium vs Low, High vs Medium) up or down by the *same amount*
* linearity of effects on this log-odds scale

## Alternative links

* logit (log-odds): *proportional odds*
* probit: very similar. Interpretable as shifting the distribution of a Gaussian latent variable

## Diagnostics

`DHARMa`, `performance::check_model()` Seem to be terrible ... ? @greenwellResiduals2018

```{r}
set.seed(101)  # for reproducibility
sres <- resids(house.plr)
p1 <- autoplot.polr(house.plr, what = "covariate", x = housing$Type)
p2 <- autoplot.polr(house.plr, what = "qq")
p1 + p2
```

May not work with `clmm`, `clmm2` (mixed models)?

[Frank Harrell on evaluating PO assumptions](https://www.fharrell.com/post/impactpo/)

## other packages

* `rms::orm`
* for mixed models: `ordinal` (`?residuals.lrm`)
* Bayesian mixed models: `brms`, `MCMCglmm`, ...

## ordinal package

* works with `emmeans`, `effects`, etc.
* testing proportionality: can set effects of some variables to be *nominal* (more flexible, allow non-proportionality) with `nominal = ~ ...`

## Example

```{r}
fit2 <- clm(Sat ~ Infl + Type + Cont, weights = Freq, data = housing)
summary(fit2)
predict(fit2, newdata = pframe, se.fit = TRUE) ## probs
```

[Frank Harrell on PO models](https://www.fharrell.com/post/rpo/)



## References

