---
title: "Bio 708/theobio style guide"
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography: "qmee.bib"
---

Follow Hadley Wickham's [style guide](http://adv-r.had.co.nz/Style.html), adapted from [Google's style guide](https://google.github.io/styleguide/Rguide.xml) (HW's link to Google is broken). Some of the differences between Google and HW are:

- HW prefers `snake_case` for identifiers, Google likes kebab or camel (`variable.name` (or `variableName`), `FunctionName`, `kConstantName`)
- HW uses Roxygen

You can use whatever naming conventions you want but **be consistent**.

In addition:

- don't use absolute paths (i.e. `C:\\Joe's Computer\\Important Stuff`)
- don't use file names with spaces or parentheses (or other special characters such as `&`)
- don't depend on case sensitivity for anything important
- don't put `rm(list=ls())` at the top of your code [@bryan_project-oriented_2017]
- always use `TRUE` and `FALSE` rather than `T`/`F` (is this in the other style guides already???)
- loading packages:
    - use `library()` instead of `require()` to load packages
	- load all of the packages you're going to use in a script at the top of the script
	- don't use `install.packages()` in a script (unless it's commented out)
- surround multi-line expressions with parentheses instead of relying on R's rule that statements should be carried on to the next line when the line ends with a binary operator (usually `+`, `%>%`). For example,
```{r cont1,eval=FALSE}
thing <- (thing %>%
          mutate(foo=x^2)
)
```
rather than
```{r cont2,eval=FALSE}
thing <- thing %>%
    mutate(foo=x^2)
```
Consider moving the operator to the next line:
```{r cont3,eval=FALSE}
thing <- (thing
     %>% mutate(foo=x^2)
)
```
This makes it easier to comment out unwanted lines temporarily.
- Similarly, for complicated multi-argument expressions, put the comma on the following line to make commenting/deleting arguments easier (JD)
```{r eval=FALSE} 
thing <- (thing 
    %>% mutate(foo=x^2
           , bar=x^3
           , bletch=x^4
	)
)
```
rather than
```{r eval=FALSE}
thing <- thing %>%
    mutate(foo=x^2,
           bar=x^3,
           bletch=x^4)
```
- use explicit `print()` statements rather than relying on objects to self-print
- use explicit `return()` statements rather than relying on R's implicit "return value is the last statement in the function" rule
- use ESS conventions of `#` for end-of-line comments, `##` for comments on a line by themselves
- omit unnecessary `c()` (e.g. `c(1:30)`). Lean toward `seq()` and `seq_along()`, but OK (?) to use `:`

## References
