---
title: "Canadian homicide rates"
author: Ben Bolker
---

```{r opts,echo=FALSE}
library(knitr)
opts_chunk$set(results="hide")  ## match wiki behaviour - don't show results
```


## Read data in wide format and play with it

Data are originally from [StatCan](http://www.statcan.gc.ca/tables-tableaux/sum-som/l01/cst01/legal12b-eng.htm) (it might be possible to *scrape* this data set using the `XML` package ...).

I also got [data on population size in 2011 from Wikipedia](http://en.wikipedia.org/wiki/List_of_Canadian_provinces_and_territories_by_population) .

```{r getdat}
dat <- read.csv("CA_homicide.csv",check.names=FALSE)
popdat <- read.csv("popdat.csv")
```
We use `check.names=FALSE` to stop R from trying to sanitize the column names: this is reasonable if we plan to convert to long form and want to preserve the values (years, in this case).

These data are in wide format:
```{r showdat}
head(dat)
```

What if we want combine other information?
```{r regiondat}
head(rdat <- data.frame(Place=dat$Place,
      Region=c("all",rep("Atlantic",4),
             rep("East",2),
             rep("West",4),
             rep("North",3))))
```

Let's start by converting the data to long form:
```{r gather}
library("tidyr")
library("dplyr")
sdat <- dat %>% gather(year,homicides,-Place,convert=TRUE)
```
(we use `convert=TRUE` to convert the years back to numeric values)

```{r join}
sdat2 <- sdat %>% full_join(rdat) %>% full_join(popdat)
```

If we just used the original data set (without the added stuff), it's fairly easy to get summary statistics by dropping the first row (so that we have a data frame that is all numeric) and computing means of rows and columns:
```{r sum1}
dmat <- dat[,-1]
rownames(dmat) <- dat[,1]
rowMeans(dmat)  ## means by place
colMeans(dmat)  ## means by year
```
(Don't forget the `na.rm` argument, unnecessary in this case, that can be provided to most R summary functions to get them to ignore `NA` values.)

If we want summary statistics from the full data set we can do

```{r,results="hide"}
sdat2 %>% group_by(Place) %>% summarise(mean=mean(homicides))
sdat2 %>% group_by(year) %>% summarise(mean=mean(homicides))
```

One more useful technique is reordering factors (representing categorical variables) in a sensible way.  Right now the 'places' (provinces, territories, etc.) are ordered alphabetically, R's default.

```{r}
sdat3 <- sdat2 %>% mutate(Place=reorder(Place,Pop_2011))
```

I can also group by two different variables:
```{r}
sdat2 %>% group_by(year,Region) %>% summarise(mean=mean(homicides))
```

What if I want the mean and standard error?  R doesn't have a built-in "standard error of the mean" function so I define one on the fly:

```{r}
sem <- function(x) { sd(x)/sqrt(length(x))}
sdat2 %>% group_by(year,Region) %>%
    summarise(mean=mean(homicides,na.rm=TRUE),
              sem=sem(homicides))
```

What if I want to check the variables to see why they're `NA`?

```{r}
sdat2 %>% filter(year==2007 & Region=="all")
```

## Hadleyverse 1

```{r hv1pkgs}
detach("package:dplyr")
detach("package:tidyr")
library("plyr")
```

We can use `merge` to combine the data (without getting tricky, we can't `merge` more than two data sets at a time)
```{r}
dat2 <- merge(dat,rdat)
dat3 <- merge(dat2,popdat)
head(dat3)
```

```{r}
## Long format
```{r reshape,message=FALSE}
library("reshape2")
```

Reshape data from wide to long ("melt"): since I added the population data, I can't rely on `melt`'s default rule (ID variables=factor variables) but have to specify `id.var` explicitly.
```{r melt1}
mdat <- melt(dat3,variable.name="Year",id.var=c("Place","Region","Pop_2011"))
```
Now clean up, relabeling the value column and translating years back to a numeric variable:
```{r meltclean}
mdat <- rename(mdat,c(value="Hom_rate"))
mdat$Year <- as.numeric(as.character(mdat$Year))
```

We could also have `melt`ed the original data set (without region or population data) and waited until we had the data in long format to merge the extra information: `merge` is pretty smart.

## Casting
How do we summarize the data if we have it in long format?

Mean by place (I have to be a little careful with the way I `rename`: since `(all)` is not a legal R name, I have to surround it with back-quotes):
```{r meanbyplace}
place_mean <- dcast(mdat,Place~.,
      value="Hom_rate",fun.aggregate=mean)
place_mean <- rename(place_mean,c(`(all)`="mean_hom"))
print(place_mean)
```

By year:
```{r meanbyyear,results="hide"}
dcast(mdat,Year~.,
      value="Hom_rate",fun.aggregate=mean)
```

By region:
```{r meanbyregion,results="hide"}
dcast(mdat,Region~.,
      value="Hom_rate",fun.aggregate=mean)
```

By region *and* year:
```{r meanbyregionyear,results="hide"}
dcast(mdat,Region+Year~.,
      value="Hom_rate",fun.aggregate=mean)
```

To get multiple summaries in the same data frame, I need `plyr::ddply`.
```{r sem}
ddply(mdat,"Region",summarise,
      mean=mean(Hom_rate),sem=sem(Hom_rate))
```

We could also cast it back into the original wide format, or into the transposed form, using `dcast(mdat,Year~Place)` or `dcast(mdat,Place~Year)`.  `reshape2` can do considerably more complicated things, too.

In the long run it is generally easier to keep your data in long format and cast it to wide as necessary.

Save the data frame we've created:

```{r save}
save("mdat",file="CA_homicide.RData")
```
